---
title: "NeotomaR Example"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_float: true
author: David Fastovich

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Accessing the Neotoma Paleoecology Database from R

Neotoma has a robust application programming interface (API) that is tightly integrated into the `neotoma` R package. This API makes calls to the constituent Neotoma databases and returns JSON files, enabling API queries in other languages like Python or even shell scripts (e.g. `curl`, `wget`). Below are links to the relevant API documentation and two examples for calling all metadata for a single site and then retreiving all ecological data from that site.

[API Documentation](https://api.neotomadb.org/api-docs/)

[Devils Lake, WI Metadata](https://api-dev.neotomadb.org/v1.5/data/sites/666) where the `666` corresponds to the unique Site ID

[Devils Lake, WI Fossil Pollen Abundance](https://api-dev.neotomadb.org/v1.5/data/downloads/684) where the `684` corresponds to the unique Dataset ID for this fossil-pollen abundance data table.


```{r message=FALSE, warning=FALSE}
## If a package is installed, it will be loaded. If any 
## are not, the missing package(s) will be installed 
## from CRAN and then loaded.

## First specify the packages of interest
packages = c("neotoma", "ggplot2", "leaflet", "reshape2", "Bchron")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```

## Querying a single site using the `neotoma` R Package

This walkthrough provides a primer on the data structures returned by `neotoma`, namely data.tables and nested lists.

```{r message=FALSE}
# Make a call for all sites with Devils Lake in their names - RETURNS RELEVANT SITE IDS
devils <- get_site(sitename = "Devils Lake")

# List all unique Site IDs - only interested in Devil's Lake, WI
# The general structure of most objected returned by the neotoma R package is either
# data.frames or nested lists.
# 
# Print all Site ID's and get site 666
devils$site.id

# Download all data for site 666 - row 1 contains the same fossil-pollen data we queried above
devils_dataset_id <- get_dataset(devils[1,]) # RETURNS RELEVANT DATASET IDS IN A NESTED LIST

# Now with the Dataset IDs in hand, lets download fossil-pollen data for Devils Lake, WI and plot
devils_pollen <- get_download(devils_dataset_id[[1]])

# Plot fossil-pollen abundances using simple function that returns ggplot object
plot_dataset <- function(dataset){
  
  # Convert to proportions
  dataset_proportion <- as.data.frame(dataset$counts/rowSums(dataset$counts))
  
  # Keep taxa that have a minimum average abundance > 5%
  dataset_proportion <- dataset_proportion[,colMeans(dataset_proportion) > .05]
  
  # Attach ages
  dataset_proportion$age <- unlist(dataset$sample.meta$age)
  
  # Converting to a narrow data format
  dataset_melt <- melt(dataset_proportion, id.vars = "age")
  
  poll_plot <- ggplot(data = dataset_melt, aes(x = age, y = value)) +
    geom_line() + 
    scale_x_reverse() + 
    facet_wrap(~variable, scales = "free_y", nrow = ncol(dataset_proportion))
  
  return(poll_plot + theme_bw())
}

plot_dataset(devils_pollen[[1]])
```

## Running an age model from chronological control data

The `get_chroncontrol` returns a nested list of age controls for a specified dataset. Let's use this funciton on the dataset of fossil-pollen abundances that we just retrieved and visualized. 

```{r message=FALSE, results='hide'}
# Returns uncalibrated radiocarbon dates
devils_chron_ctrl <- get_chroncontrol(devils_dataset_id[[1]])

# Use BChron to create a new chronology
new_chron <- Bchronology(ages   = devils_chron_ctrl[[1]]$chron.control$age, # Uncalibrated ages
                         ageSds = devils_chron_ctrl[[1]]$chron.control$age - devils_chron_ctrl[[1]]$chron.control$age.young, # Calculating error as the difference from `age.young`
                         positions = devils_chron_ctrl[[1]]$chron.control$depth, # Depth
                         positionThicknesses = devils_chron_ctrl[[1]]$chron.control$thickness, # Thickness of sample
                         calCurves = c(rep("intcal20", length(devils_chron_ctrl[[1]]$chron.control$age) - 2), "normal", "normal")) # Calibration to apply

plot(new_chron)
```

## Querying all sites with data between a time interval

Here, we will query Neotoma to return all sites in eastern North America with fossil-pollen data from 20,000 ka to 1,000 ka. With all sites retrieved, we can place the same process that we followed for Devils Lake above in a loop to return each fossil-pollen record, plot it, and rebuild the age-depth model. It takes a *very* long time to download 658 sites worth of data, so I will leave this as a leaflet plot to visualize all sites retrieved.

```{r message=FALSE}
# Make API call
ena_sites <- get_dataset(datasettype = "pollen", # Data type
                         loc = c(-92, 20, -50, 60), # Define bounding box
                         ageyoung = 1000, # Minimum age for search
                         ageold = 20000) # Maximum age for search

# Function to make leaflet plot - from Simon Goring
# http://www.goring.org/resources/neotoma_lab_meeting.html
leaflet_map <- function(dataset_in) {
  dataset_summary <- do.call(rbind, lapply(dataset_in, 
                        function(x){
                          # here we pull out the site information from the `dataset` objects:
                          data.frame(name = x$site.data$site.name,
                                     lat  = x$site.data$lat + runif(1, -0.005, 0.005),
                                     long = x$site.data$long + runif(1, -0.005, 0.005),
                                     type = x$dataset.meta$dataset.type)
                        }))
  
  # The leaflet package documentation uses piping.  For the sake of this tutorial, I won't.
  # First, define a color palette for the dataset type symbol plotting.
  pal <- leaflet::colorFactor("Dark2", domain = levels(dataset_summary$type))
  
  # Now make the leaflet map, add base raster tiles and then add the markers for the records:
  map <- leaflet(data = dataset_summary)
  map <- leaflet::addTiles(map)
  map <- leaflet::addCircleMarkers(map, ~long, ~lat, 
                                   popup = ~paste0("Site: ", as.character(name), "<br>",
                                                   "Type: ", 
                                                   as.character(dataset_summary$type)),
                                   color = ~pal(type),
                                   stroke = FALSE, fillOpacity = 0.5)
  # You need to explicitly call the `map` object to make it appear!
  map
}

leaflet_map(ena_sites)
```
